Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-23 paper/paper.tex
fix merge conflicts, delete redundant axioms

\steven{The axioms that I think we need:

@@

\begin{gather}
  c \lto c \cong I \\
  c \lto c' \cong 0, c \neq c' \\
  c \lto I \cong 0 \\
  c \lto g \otimes h \cong \left( (c \lto g) \otimes h \right) \oplus \left( (g \amp I) \otimes (c \lto h) \right) \\
  I \lto c \cong c \\
  c \lto (g \oplus h) \cong (c \lto g) \oplus (c \lto h) \\
  c \lto (c' \lto g) \cong (c' \otimes c) \lto g
\end{gather}

For point $(5)$, Brzozowski has $I \lto g \cong g$ for all grammars $g$, but I don't think that holds for us because we want to preserve ambiguity. We could instead have something like $I \lto g \cong (g \amp I)^{*} \otimes g$. For our purposes the character presentation seems sufficient, as the derivative reasoning Brzozowski uses only takes derivatives with respect to strings.

We want $c \lto g^{*} \cong (g \amp I)^{*} \otimes (c \lto g) \otimes g^{*}$, but this is derivable utilizing axiom $(4)$ and $g^{*} \cong I \oplus (g \otimes g^{*})$.

To make the DFA proof go through I think we additionally need the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso. Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,

@@

}
\begin{align*}
  c\tol c &\equiv I\\
  c\tol d &\equiv 0\\
  c\tol I &\equiv 0\\
  c\tol 0 &\equiv 0\\
  c\tol (A \otimes B) &\equiv (c\tol A) \otimes B + \nu(A) \otimes (c\tol B)\\
  c\tol A^* &\equiv (c\tol A) \otimes A^*\\
  c\tol (A + B) &\equiv (c\tol A) + (c\tol B)
\end{align*}

\pedro{The type family $\nu$ above is equal to $I$ if its argument
  parses the empty string and zero otherwise. In order to define
  it we should probably add an extra axiom, right?}

Using the $G$ modality, we can internalize some of the external
reasoning about grammars into the language of non-linear types. Recall
that the denotation of $G A$ is the set of parses of the empty
string. As is well known in formal language theory, all the parsing
behavior of a grammar can be reduced to reasoning about parsing of the
empty string by using the derivative of a grammar. For instance, the
set of parses for a fixed string $w$ can be defined as $G(w \tol A)$
and so the type of all parses of a grammar can be defined as
$\Sigma(w:\textrm{String}) G(w \tol A)$. This can be used to define
internally a notion of \emph{ambiguity} of a grammar. A grammar $A$ is
ambiguous \steven{This is an unambiguous grammar, right?} if for every string $w$, the type $G(w \tol A)$ of
$w-$parses is a \emph{proposition} in the sense that all terms in $G(w
\tol A)$ are equal.\max{I think we should add a propositional
  truncation to the type theory.}  Similarly, the type of \emph{total
parsers} can be defined as $G(\top \tol A)$. A total parser encodes a
function that selects a parse for every possible string. A
\emph{semi-parser} is then defined as $G(\top \tol A \oplus \top)$,
where the parser is allowed to fail for any reason. This has the
downside that the always failing semi-parser $\lambda _. \textrm{in}_r
()$ is valid for any grammar. If we were working with a full
\emph{bunched} type theory, we would be able to express the more
precise type of a \emph{deciding} parser as $G(\top \tol A \oplus \neg
A)$ where $\neg A = A \Rightarrow 0$ uses the intuitionistic
implication.

@@

\steven{An aside on some axioms}

Something like $A \cong (I \& A) \oplus \sum_{c} c \otimes (c \lto A)$

Also add the axiom that $\top \cong (\sum_{c \in \mathcal{A}} c)^{*}$

If we have the above, we can then show the following isos for some alphabet $\mathcal{A}$,
\begin{align*}
  A
  & \cong \top \amp A \\
  & \cong (\Sigma_{c \in \mathcal{A}})^{*}\amp A \\
  & \cong (I \oplus (\Sigma_{c} c \otimes \top)) \amp A \\
  & \cong (I \amp A) \oplus ((\Sigma_{c} c \otimes \top)) \amp A) \\
  & \cong (I \amp A) \oplus (\Sigma_{c }c \otimes (c \lto A)) & \text{using } \phi^{-1}
\end{align*}

From here we need to also show that derivatives of regular languages are regular (look at Brzozowski's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?). Then the last iso in the chain above can be used to build a DFA equivalent to $A$ using Brzozowski's algorithm.

Need to also add in axioms for $\amp$ distributing over $\oplus$ and $\lto$ distributing over $\oplus$. And that $c \amp c' \vdash 0$ for $c \neq c'$

@@

\steven{Put powerset construction here}

>>>>

\begin{align*}
  c\tol c &\equiv I\\
  c\tol d &\equiv 0\\
  c\tol I &\equiv 0\\
  c\tol 0 &\equiv 0\\
  c\tol (A \otimes B) &\equiv (c\tol A) \otimes B + \nu(A) \otimes (c\tol B)\\
  c\tol A^* &\equiv (c\tol A) \otimes A^*\\
  c\tol (A + B) &\equiv (c\tol A) + (c\tol B)
\end{align*}

\pedro{The type family $\nu$ above is equal to $I$ if its argument
  parses the empty string and zero otherwise. In order to define
  it we should probably add an extra axiom, right?}
\steven{$\nu$ can be derived using $\amp$, $\nu(A) = I \amp A$}

\steven{In addition to Pedro's axioms:

@@

\begin{gather}
  I \lto c \cong c \\
  c \lto (c' \lto g) \cong (c' \otimes c) \lto g
\end{gather}

For axiom $(1)$, Brzozowski has $I \lto g \cong g$ for all grammars $g$, but I don't think that holds for us because we want to preserve ambiguity. We could instead have something like $I \lto g \cong (g \amp I)^{*} \otimes g$. For our purposes the character presentation seems sufficient, as the derivative reasoning Brzozowski uses only takes derivatives with respect to strings.

Axiom $(2)$ might be admissible, but I'm unsure.

I think for Kleene star we need $c \lto g^{*} \cong (g \amp I)^{*} \otimes (c \lto g) \otimes g^{*}$. This should derivable utilizing the product axiom and $g^{*} \cong I \oplus (g \otimes g^{*})$, or is that sort of isomorphism opaque to us right now.

To make the DFA proof go through I think we additionally need to axiomatically make the map $c \otimes (c \lto A) \vdash (c \otimes \top) \amp A$ to be an iso. Specifically, this term needs an iso $\phi : c \otimes (c \lto A) \to (c \otimes \top) \amp A$ defined below,

@@

but this last bit isn't just Brzozowski's definitions

}
Using the $G$ modality, we can internalize some of the external
reasoning about grammars into the language of non-linear types. Recall
that the denotation of $G A$ is the set of parses of the empty
string. As is well known in formal language theory, all the parsing
behavior of a grammar can be reduced to reasoning about parsing of the
empty string by using the derivative of a grammar. For instance, the
set of parses for a fixed string $w$ can be defined as $G(w \tol A)$
and so the type of all parses of a grammar can be defined as
$\Sigma(w:\textrm{String}) G(w \tol A)$. This can be used to define
internally a notion of \emph{ambiguity} of a grammar. A grammar $A$ is
ambiguous \steven{This is an unambiguous grammar, right?} if for every string $w$, the type $G(w \tol A)$ of
$w-$parses is a \emph{proposition} in the sense that all terms in $G(w
\tol A)$ are equal.\max{I think we should add a propositional
  truncation to the type theory.}  Similarly, the type of \emph{total
parsers} can be defined as $G(\top \tol A)$. A total parser encodes a
function that selects a parse for every possible string. A
\emph{semi-parser} is then defined as $G(\top \tol A \oplus \top)$,
where the parser is allowed to fail for any reason. This has the
downside that the always failing semi-parser $\lambda _. \textrm{in}_r
()$ is valid for any grammar. If we were working with a full
\emph{bunched} type theory, we would be able to express the more
precise type of a \emph{deciding} parser as $G(\top \tol A \oplus \neg
A)$ where $\neg A = A \Rightarrow 0$ uses the intuitionistic
implication.

@@

% \steven{An aside on some axioms}

% Something like $A \cong (I \& A) \oplus \sum_{c} c \otimes (c \lto A)$

% Also add the axiom that $\top \cong (\sum_{c \in \mathcal{A}} c)^{*}$

% If we have the above, we can then show the following isos for some alphabet $\mathcal{A}$,
% \begin{align*}
%   A
%   & \cong \top \amp A \\
%   & \cong (\Sigma_{c \in \mathcal{A}})^{*}\amp A \\
%   & \cong (I \oplus (\Sigma_{c} c \otimes \top)) \amp A \\
%   & \cong (I \amp A) \oplus ((\Sigma_{c} c \otimes \top)) \amp A) \\
%   & \cong (I \amp A) \oplus (\Sigma_{c }c \otimes (c \lto A)) & \text{using } \phi^{-1}
% \end{align*}

% From here we need to also show that derivatives of regular languages are regular (look at Brzozowski's proof) and that regular languages have finitely many derivatives (potentially with some size constraint on the derivative?). Then the last iso in the chain above can be used to build a DFA equivalent to $A$ using Brzozowski's algorithm.

% Need to also add in axioms for $\amp$ distributing over $\oplus$ and $\lto$ distributing over $\oplus$. And that $c \amp c' \vdash 0$ for $c \neq c'$

@@
